 

### 2021-05-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Inherent leaf $gs$ is an important factor for $T_{leaf}$ which is typically greater across species exposed sun-leaves than buffered shade leaves. However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress.

@@

Across forest species widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds due to inherent specie-specific traits. Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007]. Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Therefore, in inter-specific trait variation, while general patterns of thermoregulation can be observed, some species might be more suspectable to heat stress at a vertical gradient than others [@fauset_differences_2018].

@@

##### *Leaf lifespan* 
*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

Deciduousness as a leaf trait is an adaptation to water and heat stress. Therefore, can influence biophysical gradient and $T_{leaf}$ distribution during leaf-loss [@rey-sanchez_spatial_2016]. *In the tropics, among species that can be deciduous, greater proportion of deciduous individuals are in larger size classes that tend to lose their leaves more than understory species that are seldom deciduous. Deciduous species are also more prominent in drier sites compared to wet sites [@conditQuantifyingDeciduousnessTropical2000b]*

*Spring and fall leaf phenology in temperate deciduous forests (Augsburger).* Leaf emergence in deciduous forests is delayed with height-ontogeny where understory species leaf-out earlier than canopy trees [Vitasse et al. 2013; @seiwaChangesLeafPhenology1999].

@@

### Stomatal conductance

While exposed canopy sun-leaves are equipped to thermoregulate more efficiently than shaded-leaves, they are largely constrained by $gs$ [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]; and therefore their ability to photosynthesize as well. During the day, max transpiration rate and $gs$ increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010]. During midday, stomtal closure in exposed leaves increases $T_{leaf}$, and decreases photosynthesis [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].

@@

**Photosynthesis is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun leaves. Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P (and water availability).

$T_{leaf}$ beyond photosynthetic $T_{opt}$, declines photosynthesis due it's peaked response to temperature. Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g. Sage & Kubien 2007; Varhammer et al. 2015 ). Heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020]. Leaves in tropical upper canopies have shown to experience in-situ $T_{air}$ well above $40-60{^\circ}C$ on a regular basis. If $T_{leaf}$ is not regulated or adapated, heat stress can loss of canopy leaves [**REF**]

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020].Similar results were shown for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].

@@

**Leaf thermal tolerance (Tcrit/ T50)**
$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [Slot et al. 2020]. Leaf thermal tolerances strongly reflect species adaptation to their microenvironment rather than macroclimate conditions [@feeley_thermal_2020; Slot et al. 2020; Perez and Feely 2020a] and phylogeny [Slot et al. 2020; Perez and Feeley 2020a].

Across latitude and climate, typical $T_{50}$ values are ~45-50$^\circ$C (O' sullivan et al. 2017). With elevation, $T_{50}$ and $T_{crit}$ decrease, as seen in tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], and vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020]. Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$. On the other hand, leaves of species with greater construction costs--greater LMA-- had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020]. However, higher photosynthetic heat tolerance does not necessarily increase the breadth of leaf thermal safety margin. Rather, @perezPhotosyntheticHeatTolerances2020 found that, across species sun leaves, higher photosynthetic heat tolerance experienced higher $T_{leaf_{max}}$, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf}$-$T_{50}$)

**Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
$T_{crit}$ and $T_{50}$ over the years have been predominantly measured on sun leaves, hence there is limited information on sun- and shade-leaves  and at vertical height gradient.

Knapp et al. 1994 showed that in tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions. In both species photosynthetic heat tolerance increased seasonally from June to August. However, more recetly @slotPhotosyntheticHeatTolerance2019a found that though $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Paname tropical forest, the results weren't significant, with both measured near ground level.

In other cases, leaf thermal tolerance reflects microclimatic stress along a vertical gradient. For *Acacia Papyrocarpa* (Benth.) trees in an arid region in Australia, higher composite climate stress and indeed, $T_{50}$ was greater in the lower north-facing canopy than other canopy positions. Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, low-wind speed and greater $T_{air}$ [@curtis_intracanopy_2019].

@@

Across species in north-eastern oak deciduous forest, conifer temperate rainforest and in *Liquidambara styraciflua* tree, $Q_{10}$ is observed to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996]. However, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig. 1*)[@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.

@@

Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al. 2016]. On the other hand, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in canopy tops where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. However, soil water loss in conditions of drought effect larger trees more than small trees. Primarily because, limited soil-water cannot meet long-distance water transport to the canopy's biophysical evaporative demand [Norway spruce study, bennett; @mcgregor_tree_nodate].

On the basis of phenology, because understory trees leaf out earlier than canopy [Vitasse et al. 2013], understory growth is more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate-deciduous forests [Zohnner and Renner, 2019]. However, late spring short-term heat waves negatively impact partially- expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis in reflushed leaves, more than understory saplings that are buffered from such conditions [Filewood and Thomas, 2013].

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).
It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. [Increase in humidity?]

Additionally, interacting biophysical and biological stressors show that stress sensitivity is greater in buffered small trees, and decreases where tolerance increases with height, light with available water.  Partly because large NSC reserves in taller trees due to available light allow them to tolerate greater number of stressors than buffered trees that are light and NSC limited [@niinemetsResponsesForestTrees2010a]

@@

Canopy structural composition and $LAI$ largely influence understory species composition and soil fertility [@majasalmiImpactTreeCanopy2020], buffering capacity, which in-turn is influenced by the hydrological cycle [@davis_microclimatic_2019]. Each of these are subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a]. Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and decreases tree canopy circumference [@aussenacInteractionsForestStands2000a]. With larger trees suffering disproportionally greater hydraulic stress during drought, can contribute to increasing larger tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020], or increase in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness, also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].

Increasing global temperatures can increase canopy $T_{leaf}$ sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].  Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020]. Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020]. Each of these changes have potential feedbacks to climate change.

@@

While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]. Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020]. Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020]. Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to microclimatic fauna [@scheffersMicrohabitatsReduceAnimal2013] . Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020]. Whereas warming in other cases is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]. Evidence shows that warming and heat waves can also drastically alter canopy and understory phenology and forest photosynthesis [@zohnerOngoingSeasonallyUneven2019].

Additionally, warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth). Differential mortality within the canopy structure will alter the profile of metabolism and hence emergent forest function. Such alteration to canopy structure may be detectable by lidar and thermal remote sensing. If we know how metabolism maps to the thermal profile, then the observed forest structural changes can inform prediction of forest function.

>>>>

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Inherent leaf $gs$ is an important factor for $T_{leaf}$ which is typically greater across species exposed sun-leaves due to stomata density than shade leaves. However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress.

@@

Across forest species widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds due to inherent specie-specific traits. Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007]. Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Therefore, while general patterns of thermoregulation can be observed across species, some might be more suspectable to heat stress at a vertical gradient than others [@fauset_differences_2018].

@@

##### *Leaf lifespan*

New-leaf growth across the vertical gradient towards light constantly shades leaves underneath. Therefore, exposed-canopies have greater proportion of young leaves than shaded regions that have older leaves. Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger leaves that are devoloping traits than in mature leaves where traits have been developed. In such cases, change in light can result in dropping of leaves [Niinemetes, 2007]. *Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

Deciduousness as a leaf trait is an adaptation to water and heat stress. Indeed, greater proportion of deciduous species are found in drier tropicaal regions than wetter,  and among species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000b]. Therefore, deciduous leaf-loss can influence biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016]. Spatio-temporal leaf-emergence in deciduous-temperate forests also shapes biophysical gradient where leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al. 2013; @seiwaChangesLeafPhenology1999].

@@

### Stomatal conductance [$gs]

While $Gs$ mediates crucial leaf-atmosphere interaction across species, their exposed-sun-leaves, though equipped to thermoregulate efficiently are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]. Max transpiration rate and $gs$ is peaks in exposed-canopies during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines through midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].

@@

**Photosynthesis is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun-leaves. Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.

In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P (and water availability).

$T_{leaf}$ beyond photosynthetic $T_{opt}$, declines photosynthesis due it's peaked response to temperature. Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g. Sage & Kubien 2007; Varhammer et al. 2015 ). Heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies that compared temperature sensitivity of photosynthesis in sun and shade leaves.

For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020].Similar results were shown for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun- and shade-leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].

**Leaf thermal tolerance (Tcrit/ T50)**

$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [Slot et al. 2020].

**Thermal tolerance (e.g., $T_{50}$) sparsely varies with exposure among recent studies, but there are no studies isolating effect of height.**

150 years ago,  Sachs studied garden species $T_{50}$. Though, with little information on shade-leaves [*REF*]. In 1994, Knapp et al. showed that in tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions. In both species photosynthetic heat tolerance increased seasonally from June to August. However, more recetly @slotPhotosyntheticHeatTolerance2019a found that though $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Panama tropical forest, the results weren't significant, with both measured near ground level.

$T_{50}$ in *Acacia Papyrocarpa* (Benth.) trees in an arid region was observed to reflect microclimatic stress index along a vertical gradient where lower north-facing canopy showed higher composite climate stress and greater $T_{50}$ than other canopy positions, that correlated with low wind-speed greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].

@@

Recent studies show that while leaf thermal tolerances across species reflects local microenvironmental conditions than regional macroclimate, [@feeley_thermal_2020; Slot et al. 2020; Perez and Feely 2020a] $T_{50}$ does not vary with phylogeny [Slot et al. 2020; Perez and Feeley 2020a].

Global $T_{50}$s across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al. 2017). Additionally, $T_{50}$ and $T_{crit}$, decrease with elevation in tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020]. $T_{crit}$ is also recently found to be greater in species with large leaves, and slow heating and cooling time [higher thermal time constant] and $T_{50}$ is found to be greater among species' leaves with greater construction costs--greater LMA--compared to species with low LMA [@slotLeafHeatTolerance2020]. @perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance experienced higher $T_{leaf_{max}}$, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$).

@@

Across species in north-eastern oak deciduous forest, conifer temperate rainforest and in *Liquidambara styraciflua* tree, $Q_{10}$ is observed to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996]. In contrast, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig. 1*)[@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.

@@

Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al. 2016].This can implication to canopies and gaps that reside at levels of greater $VPD$. Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. However, soil water loss during drought effects larger trees more than small trees because, limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canonopy [Norway spruce study, bennett; @mcgregor_tree_nodate].

Additionally, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate-deciduous forests [Zohnner and Renner, 2019]. However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves by more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis [and C uptake?] in reflushed leaves [Filewood and Thomas, 2013].

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).
It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. Interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water. Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].

@@

The complex interactions of the biophysical environment and biological--leaf traits, metabolic processes and distribution of trees--are largely interconneced, resulting in feedbacks to another. Therefore, climate-driven changes, such as rising temperature, can have implications to forests at several scales.

Canopy buffering capacity which is influenced by hydrological-cycle at large is crucial to microclimates [Davis et al. 2019], where disturbance to canopy structural composition and $LAI$ influences understory species composition, soil fertility [*@majasalmiImpactTreeCanopy2020*] and species in microclimate conditions, particularly threatening flora [REF] fauna adapted to microrefugias [REF]. Such disturbances to canopies are observed, in part, by frequent droughts, where less severe drought can decrease tree canopy circumference [@aussenacInteractionsForestStands2000a], and more severe droughts can contribute to increasing larger tree mortality in forests around the world [@senfCanopyMortalityHas2018a; @mcdowellMechanismsPlantSurvival2008; Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Hydraulic limitation in forests can also alters species thermoregulation [@sastry_leaf_2018]. Increasing temperatures can increase canopy $T_{leaf}$ sometimes above ambient temperature, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].  Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020]. Recent adavances in canopy studies point toward large loss of canopies in Europe caused by wind-induced disturbance and anthropogenic deforestation [Cornelius Senf, 2021; @senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020]. Loss of big trees (e.g.@senfCanopyMortalityHas2018a ) would increase canopy roughness, and reductions in canopy cover also associated with reduced canopy structural complexity, can lead to altered biophysical gradient and microclimates with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020]. Each of these changes have potential feedbacks to climate change.

@@

While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020] leading to understory plant communities already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to microclimatic fauna [@scheffersMicrohabitatsReduceAnimal2013] . Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020]. Whereas warming in other cases is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]. Warming and heat waves can also drastically alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].

warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth). Differential mortality within the canopy structure will alter the profile of metabolism and hence emergent forest function. Such alteration to canopy structure may be detectable by lidar and thermal remote sensing. If we know how metabolism maps to the thermal profile, then the observed forest structural changes can inform prediction of forest function.
